A comparison of design-based and model-based approaches for finite population spatial data.

### 2022-04-13 inst/manuscript/manuscript.tex
review edits through 2.1

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data. In the simulated data, a variety of sample sizes, location
    layouts, dependence structures, and response types are considered. In
    the simulated and real data, the population mean is the parameter of
    interest and performance is measured using statistics like bias,
    squared error, and interval coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the Generalized Random
    Tessellation Stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    Simple Random Sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based approaches tend to outperform design-based approaches,
    even for skewed data where the model-based distributional assumptions
    are violated. The performance gap between these approaches is small
    GRTS samples are used but large when SRS samples are used. This
    suggests that the sampling choice (whether to use GRTS or SRS) is most
    important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units -- this subset is called a sample. There are two general
approaches for using samples to make frequentist statistical inferences
about a population: design-based and model-based. In the design-based
approach, inference relies on randomly assigning some population units
to be in the sample (random sampling). Alternatively, in the model-based
approach, inference relies on distributional assumptions about the
underlying stochastic process that generated the sample. Each paradigm
has a deep historical context (Sterba, 2009) and its own set of benefits
and drawbacks (Hansen et al., 1983). In this manuscript, we compare the
design-based and model-based approaches for finite population spatial
data.

Spatial data are data that incorporate the locations of the population
units into either the sampling or estimation process. De Gruijter and
Ter Braak (1990) and Brus and DeGruijter (1993) give early comparisons
of design-based and model-based approaches for spatial data, quashing
the belief that design-based approaches could not be used for spatially
correlated data. Since then, there have been several general comparisons
between design-based and model-based approaches for spatial data (Brus
and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008). Cooper (2006)
reviews the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design-based and
model-based principles (see e.g., Sterba (2009) and Cicchitelli and
Montanari (2012), and for Bayesian approaches, see Chan-Golston et al.
(2020) and Hofman and Brus (2021)).

Certainly comparisons between design-based and model-based approaches
have been studied in spatial contexts. Our contribution is comparing
design-based approaches that incorporate spatial locations into sampling
and analysis to model-based approaches. Though the broad comparisons we
draw between design-based and model-based approaches generalize to
finite and infinite populations, we focus on finite populations. A
finite population contains a finite number of population units (we
assume the finite number is known); an example is lakes (treated as a
whole with the lake centroid representing location) in the contiguous
United States. An infinite population contains an infinite number of
population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
the design-based and model-based approaches to finite population spatial
data. In Section \ref{sec:mm}, we describe how we compare performance of
the approaches with a simulation study and an analysis of real data that
contains mercury concentration in lakes located in the contiguous United
States. In Section \ref{sec:results}, we present results from the
simulation study and the mercury concentration analysis. And in Section
\ref{sec:discussion}, we end with a discussion and provide directions
for future research.

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of population units according to a
sampling design. A sampling design assigns a probability of selection to
each sample (a subset of population units). Some examples of commonly
used sampling designs include simple random sampling, stratified random
sampling, and cluster sampling. The inclusion probability of a
population unit follows by summing each population unit's probability of
selection in each sample containing that population unit. Inclusion
probabilities are later used to estimate population parameters.

When samples are chosen in a manner such that the layout of sampled
units reflects the layout of the population units, we call the resulting
sample ``spatially balanced.'' By ``reflecting the layout of the
population units'', we mean that if population units are concentrated in
specific areas, the units in the sample should be concentrated in the
same areas. Because spatially balanced samples reflect the layout of the
population units, they are not necessarily ``spread out'' in space in
some equidistant manner. One approach to selecting spatially balanced
samples is the Generalized Random Tessellation Stratified (GRTS)
algorithm (Stevens and Olsen, 2004), which we discuss in more detail in
Section \ref{subsec:spb_design}.

@@

The model-based approach assumes the population is a random realization
of a data-generating stochastic process called a superpopulation.
Randomness is formally incorporated through distributional assumptions
on the superpopulation. Strictly speaking, randomness need not be
incorporated through random sampling, though Diggle et al. (2010) warn
against preferential sampling. Preferential sampling occurs when the
process generating the data locations and the process being modeled are
not independent of one another. To guard against preferential sampling,
model-based approaches can implement some form of random sampling,
though it is common for model-based approaches to not implement random
sampling. When model-based approaches do implement random sampling, the
inclusion probabilities are ignored when analyzing the sample (in
contrast to the design-based approach, which relies on these inclusion
probabilities to analyze the sample).

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable. For example, suppose the realized mean of
all population units (the realized population mean) is the variable of
interest. Instead of a fixed, unknown mean, we are predicting the value
of the mean, a random variable. Prediction intervals are then derived
using assumptions of the data-generating stochastic process. If we
repeatedly generate response values from the same process and select
samples, then 95\% of all 95\% prediction intervals constructed from a
procedure with appropriate coverage will contain their respective
realized means. Cressie (1993) and Schabenberger and Gotway (2017)
provide thorough reviews of model-based approaches for spatial data. In
Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based
and model-based approaches (Ver Hoef (2002) and Brus (2021) provide
similar figures).

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/dvm_comp.jpeg}
  \caption{A visual comparison of the design-based and model-based approaches. In the top row, the design-based approach is highlighted. There is one fixed population with nine population units and three random samples of size four (points circled are those sampled). The response values at each site are fixed, but we obtain different estimates for the mean response in each random sample. In the bottom row, the model-based approach is highlighted. There are three realizations of the same data-generating stochastic process that are all sampled at the same four locations. The data-generating stochastic process has a single mean, but the mean of the nine population units is different in each of the three realizations.}
  \label{fig:fig1}
\end{figure}

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less than
parameter estimates from samples that are not spatially balanced
(Barabesi and Franceschi, 2011; Benedetti et al., 2017; Grafström and
Lundström, 2013; Robertson et al., 2013; Stevens and Olsen, 2004; Wang
et al., 2013). The first spatially balanced sampling algorithm to see
widespread use was the Generalized Random Tessellation Stratified (GRTS)
algorithm (Stevens and Olsen, 2004). To quantify the spatial balance of
a sample, Stevens and Olsen (2004) proposed loss metrics based on
Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was
developed, several other spatially balanced sampling algorithms emerged,
including stratified sampling with compact geographical strata Walvoort
et al. (2010), the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we select spatially balanced samples using the
Generalized Random Tessellation Stratified (GRTS) algorithm because it
is readily available in the \texttt{spsurvey} \textbf{\textsf{R}}
package (Dumelle et al., 2022) and naturally accommodates finite and
infinite sampling frames, unequal inclusion probabilities, and
replacement units (replacement units are population units that can be
sampled when a population unit originally selected can no longer be
sampled).

The GRTS algorithm selects samples by utilizing a particular mapping
between two-dimensional and one-dimensional space that preserves
proximity relationships. First the bounding box of the domain is split
up into four distinct, equally sized squares called level-one cells.
Each level-one is randomly assigned an level-one address of 0, 1, 2, or
3. The set of level-one cells is denoted by \(\mathcal{A}_1\) and
defined as \(\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}\). Within
each level-one cell, the inclusion probability for each population unit
is summed, and if any of these sums exceed one, a second level of cells
is added. Then each level-one cell is split into four distinct, equally
sized squares called level-two cells. Each level-two cell is randomly
assigned a level-two address of 0, 1, 2, or 3. The set of level-two
cells is denoted by \(\mathcal{A}_2\) and defined as
\(\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}\).
The inclusion probabilities within each level-two cell are summed, and
if any of these sums exceed one, a third level of cells is added. This
process continues for \(k\) steps, until all level-\(k\) cells have
inclusion probability sums no larger than one. Then
\(\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}\).

After determining \(\mathcal{A}_k\), it is placed into hierarchical
order. Hierarchical order is a numeric order that first sorts
\(\mathcal{A}_k\) by the level-one addresses from smallest to largest,
then sorts \(\mathcal{A}_k\) by the level-two addresses from smallest to
largest, and so on. For example, \(\mathcal{A}_2\) in hierarchical order
is the set
\(\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}\). Because
hierarchical ordering sorts by level-one cells, then level-two cells,
and so on, population units that have similar hierarchical addresses
tend to be nearby one another in space.Next each population unit is
mapped to a one-dimensional line in hierarchical order where each
population unit's inclusion probability equals its line-length. If a
level-\(k\) cell has multiple population units in it, they are randomly
placed within the cell's respective line segment. A uniform random
variable is then simulated in \([0, 1]\) and a systematic sample is
selected on the line, yielding \(n\) sample points for a sample size
\(n\). Each element in this systematic sample falls on some population
unit's line segment, and thus that population unit is selected in the
sample. For further details regarding the GRTS algorithm, see Stevens
and Olsen (2004).

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimator for \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th population unit in
the sample, \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample, and \(n\) is the sample size. An estimate
of the population mean is obtained by dividing \(\hat{\tau}_{ht}\) by
\(N\), the number of population units.

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically
for GRTS samples. Second, these estimators tend to ignore the spatial
locations of the population units. To address these two drawbacks
simultaneously, Stevens and Olsen (2003) proposed the local neighborhood
variance estimator. The local neighborhood variance estimator does not
rely on \(\pi_{ij}\) and estimates the variance of \(\hat{\tau}\)
conditional on the random properties of the GRTS sample -- the idea
being that this conditioning should yield a more precise estimate of
\(\hat{\tau}\). They show that the each observation's contribution to
the overall variance is dominated by local variation. Thus the local
neighborhood variance estimator is a weighted sum of variance estimates
from each observation's local neighborhood. These local neighborhoods
contain observation itself and its three nearest neighbors. For more
details, see Stevens and Olsen (2003).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- for technical details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

@@

REWRITE AS SIMPLE RANDOM SAMPLING AND WITHOUT REPLACEMENT

USE SRS / GRTS - DB / MB

ADD LOHR REFERENCE AND FORM FOR SRS VARIANCE WITHOUT REPLACEMENT WITH
FPC

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
was IRS-Design. In IRS-Design, samples were selected with the
Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores
the spatial locations of the population units, thus the IRS samples were
not spatially balanced. In IRS-Design, samples were analyzed using the
design-based approach via the Horvitz-Thompson mean estimator and an IRS
variance estimator that ignored the spatial locations of the units in
the sample. The second sampling-analysis combination was IRS-Model,
where samples were selected with the IRS algorithm and analyzed using
the model-based approach while estimating the covariance parameters
(\(\bm{\delta}\)) and fixed effects (\(\bm{\beta}\) using restricted
maximum likelihood (REML). The third sampling-analysis combination was
GRTS-Design, where samples were selected with the GRTS algorithm and
analyzed using the design-based approach via the Horvitz-Thompson mean
estimator and the local neighborhood variance estimator (which does
incorporate the spatial locations of the units in the sample). The
fourth and final sampling-analysis combination was GRTS-Model, where
samples were selected with the GRTS algorithm and analyzed using the
model-based approach while estimating the covariance parameters
(\(\bm{\delta}\)) and fixed effects (\(\bm{\beta}\) using restricted
maximum likelihood (REML). These sampling-analysis combinations are also
provided in Table \ref{tab:designanalysis}. Lastly we note that for both
the IRS and GRTS samples, equal inclusion probabilities were assumed for
all population units. When IRS assumes equal inclusion probabilities for
all population units, the algorithm is equivalent to simple random
sampling (SRS).

\begin{table}[ht]
\centering
\begin{tabular}{r|ll}
  \hline
 & Design & Model \\ 
  \hline
IRS & IRS-Design & IRS-Model \\ 
  GRTS & GRTS-Design & GRTS-Model \\ 
   \hline
\end{tabular}
\caption{\label{tab:designanalysis} Sampling-analysis combinations in the simulation study. The rows give the two types of sampling designs and the columns give the two types of analyses.} 
\end{table}

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error (DRE). The three sample sizes (\(n\)) were \(n = 50, n = 100,\)
and \(n = 200\). Samples were always selected from a population size
(\(N\)) of \(N = 900\). The two location layouts were random and
gridded. Locations in the random layout were randomly generated inside
the unit square (\([0, 1] \times [0, 1]\)). Locations in the gridded
layout were placed on a fixed, equally spaced grid inside the unit
square. The two response types were normal and lognormal. For the normal
response type, the response was simulated using mean-zero random errors
with the exponential covariance (Equation\(~\)\ref{equation:expcov}) for
varying proportions of dependent random error. The proportion of
dependent random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget) from
Equation\(~\)\ref{equation:expcov}, respectively. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), chosen so that the correlation in the dependent random
error decayed to nearly zero at \(\sqrt{2}\), the largest possible
distance between two population units in the domain. For the lognormal
response type, the response was first simulated using the same approach
as for the normal response type, except that the total variance was
0.6931 instead of 2. The response was then exponentiated, yielding a
lognormal random variable whose total variance was 2. The lognormal
responses were used to evaluate performance of the sampling-analysis
approaches for data that were skewed (i.e., not normal).

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data. In the simulated and real data, a variety of sample sizes,
    location layouts, dependence structures, and response types are
    considered. The population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the generalized random
    tessellation stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    simple random sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based approaches tend to outperform design-based approaches,
    even for skewed data where the model-based distributional assumptions
    are violated. The performance gap between these approaches is small
    GRTS samples are used but large when SRS samples are used. This
    suggests that the sampling choice (whether to use GRTS or SRS) is most
    important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units -- this subset is called a sample. There are two general
approaches for using samples to make frequentist statistical inferences
about a population: design-based and model-based. In the design-based
approach, inference relies on randomly assigning some population units
to be in the sample (random sampling). Alternatively, in the model-based
approach, inference relies on distributional assumptions about the
underlying stochastic process generating the sample. Each paradigm has a
deep historical context (Sterba, 2009) and its own set of benefits and
drawbacks (Hansen et al., 1983, p. @brus1997random). In this manuscript,
we compare the design-based and model-based approaches for finite
population spatial data.

Spatial data are data that have some sort of spatial index, usually via
coordinates. De Gruijter and Ter Braak (1990) and Brus and DeGruijter
(1993) give early comparisons of design-based and model-based approaches
for spatial data, quashing the belief that design-based approaches could
not be used for spatially correlated data. Since then, there have been
several general comparisons between design-based and model-based
approaches for spatial data (Brus and De Gruijter, 1997; Brus, 2021; Ver
Hoef, 2002, 2008). Cooper (2006) reviews the two approaches in an
ecological context before introducing a ``model-assisted'' variance
estimator that combines aspects from each approach. In addition to
Cooper (2006), there has been substantial research and development into
estimators that use both design-based and model-based principles (see
e.g., Sterba (2009) and Cicchitelli and Montanari (2012), and for
Bayesian approaches, see Chan-Golston et al. (2020) and Hofman and Brus
(2021)).

While comparisons between design-based and model-based approaches have
been studied in spatial contexts, our contribution is comparing
design-based approaches specifically built for spatial data to
model-based approaches. Though the broad comparisons we draw between
design-based and model-based approaches generalize to finite and
infinite populations, we focus on finite populations. A finite
population contains a finite number of population units (we assume the
finite number is known); an example is lakes (treated as a whole with
the lake centroid representing location) in the conterminous United
States. An infinite population contains an infinite number of population
units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
design-based and model-based approaches to finite population spatial
data. In Section \ref{sec:mm}, we describe how we intend to compare
performance of the approaches using simulated and real data. In Section
\ref{sec:results}, we present analysis reslts for the simulated and real
data. And in Section \ref{sec:discussion}, we end with a discussion and
provide directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of population units according to a
sampling design. A sampling design assigns a probability of selection to
each sample (subset of population units). Some examples of commonly used
sampling designs include simple random sampling, stratified random
sampling, and cluster sampling. The inclusion probability of a
population unit follows by summing each sample's probability of
selection over all samples that contain the population unit. Inclusion
probabilities are later used to estimate population parameters.

When samples are chosen in a manner such that the layout of sampled
units reflects the layout of the population units, we call the resulting
sample spatially balanced. By ``reflecting the layout of the population
units'', we mean that if population units are concentrated in specific
areas, the units in the sample should be concentrated in the same areas.
Because spatially balanced samples reflect the layout of the population
units, they are not necessarily spread out in space in some equidistant
manner. One approach to selecting spatially balanced samples is the
generalized random tessellation stratified (GRTS) algorithm (Stevens and
Olsen, 2004), which we discuss in more detail in Section
\ref{subsec:spb_design}.

@@

The model-based approach assumes the population is a random realization
of a data-generating stochastic process. Randomness is formally
incorporated through distributional assumptions on this process.
Strictly speaking, randomness need not be incorporated through random
sampling, though Diggle et al. (2010) warn against preferential
sampling. Preferential sampling occurs when the process generating the
data locations and the process being modeled are not independent of one
another. To guard against preferential sampling, model-based approaches
can implement some form of random sampling, though it is common for
model-based approaches to sample non-randomly. When model-based
approaches do implement random sampling, the inclusion probabilities are
ignored when analyzing the sample (in contrast to the design-based
approach, which relies on these inclusion probabilities to analyze the
sample).

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable. For example, suppose the realized mean of
all population units (the realized population mean) is the variable of
interest. Instead of a fixed, unknown mean, we are predicting the value
of the mean, a random variable. Prediction intervals are then derived
using assumptions of the data-generating stochastic process. If we
repeatedly generate realizations from the same process and select
samples, then 95\% of all 95\% prediction intervals constructed from a
procedure with appropriate coverage will contain their respective
realized means. Cressie (1993) and Schabenberger and Gotway (2017)
provide thorough reviews of model-based approaches for spatial data. In
Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based
and model-based approaches (Ver Hoef (2002) and Brus (2021) provide
similar figures). This figure contrasts the design-based approach with a
fixed population and random sampling to the model-based approach with
random populations and non-random sampling.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/dvm_comp.jpeg}
  \caption{A visual comparison of the design-based and model-based approaches. In the top row, the design-based approach is highlighted. There is one fixed population with nine population units and three random samples of size four (points circled are those sampled). The response values at each site are fixed. In the bottom row, the model-based approach is highlighted. There are three realizations of the same data-generating stochastic process that are all sampled at the same four locations. The response values at each site are random.}
  \label{fig:fig1}
\end{figure}

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less than
parameter estimates from samples lacking spatial balance (Barabesi and
Franceschi, 2011; Benedetti et al., 2017; Grafström and Lundström, 2013;
Robertson et al., 2013; Stevens and Olsen, 2004; Wang et al., 2013). The
first spatially balanced sampling algorithm to see widespread use was
the generalized random tessellation stratified (GRTS) algorithm (Stevens
and Olsen, 2004). To quantify the spatial balance of a sample, Stevens
and Olsen (2004) proposed loss metrics based on Voronoi polygons (i.e.,
Dirichlet Tessellations). After the GRTS algorithm was developed,
several other spatially balanced sampling algorithms emerged, including
stratified sampling with compact geographical strata Walvoort et al.
(2010), the local pivotal method (Grafström et al., 2012; Grafström and
Matei, 2018), spatially correlated Poisson sampling (Grafström, 2012),
balanced acceptance sampling (Robertson et al., 2013),
within-sample-distance sampling (Benedetti and Piersimoni, 2017), and
Halton iterative partitioning sampling (Robertson et al., 2018). In this
manuscript, we select spatially balanced samples using the GRTS
algorithm because it is readily available in the \texttt{spsurvey}
\textbf{\textsf{R}} package (Dumelle et al., 2022) and naturally
accommodates finite and infinite sampling frames, unequal inclusion
probabilities, and replacement units. Replacement units are additional
population units that can be sampled when a population unit originally
selected can no longer be sampled. A couple reasons why an originally
selected site can no longer be sampled include its location being
physically inaccessible or on private land that the researcher does not
have permission to access.

The GRTS algorithm selects samples by utilizing a particular mapping
between two-dimensional and one-dimensional space that preserves
proximity relationships. First the bounding box of the domain is split
up into four distinct, equally sized squares called level-one cells.
Each level-one cell is randomly assigned a level-one address of 0, 1, 2,
or 3. The set of level-one cells is denoted by \(\mathcal{A}_1\) and
defined as \(\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}\). Within
each level-one cell, the inclusion probability for each population unit
is summed, and if any of these sums exceed one, a second level of cells
is added. Then each level-one cell is split into four distinct, equally
sized squares called level-two cells. Each level-two cell is randomly
assigned a level-two address of 0, 1, 2, or 3. The set of level-two
cells is denoted by \(\mathcal{A}_2\) and defined as
\(\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}\).
The inclusion probabilities within each level-two cell are summed, and
if any of these sums exceed one, a third level of cells is added. This
process continues for \(k\) steps, until all level-\(k\) cells have
inclusion probability sums no larger than one. Then
\(\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}\).

After determining \(\mathcal{A}_k\), it is placed into hierarchical
order. Hierarchical order is a numeric order that first sorts
\(\mathcal{A}_k\) by the level-one addresses from smallest to largest,
then sorts \(\mathcal{A}_k\) by the level-two addresses from smallest to
largest, and so on. For example, \(\mathcal{A}_2\) in hierarchical order
is the set
\linebreak \(\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}\).
Because hierarchical ordering sorts by level-one cells, then level-two
cells, and so on, population units that have similar hierarchical
addresses tend to be nearby one another in space. Next each population
unit is mapped to a one-dimensional line in hierarchical order where
each population unit's inclusion probability equals its line-length. If
a level-\(k\) cell has multiple population units in it, they are
randomly placed within the cell's respective line segment. A uniform
random variable is then simulated in \([0, 1]\) and a systematic sample
is selected on the line, yielding \(n\) sample points for a sample size
\(n\). Each of these sample points falls on some population unit's line
segment, and thus that population unit is selected in the sample. For
further details regarding the GRTS algorithm, see Stevens and Olsen
(2004).

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimator for \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n z_i \pi_i^{-1},
\end{align} where \(z_i\) is the value of the \(i\)th population unit in
the sample, \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample, and \(n\) is the sample size. An estimate
of the population mean is obtained by dividing \(\hat{\tau}_{ht}\) by
\(N\), the number of population units.

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically
for GRTS samples. Second, these estimators tend to ignore the spatial
locations of the population units. To address these two drawbacks
simultaneously, Stevens and Olsen (2003) proposed the local neighborhood
variance estimator. The local neighborhood variance estimator does not
rely on \(\pi_{ij}\) and estimates the variance of \(\hat{\tau}\)
conditional on the random properties of the GRTS sample -- the idea
being that this conditioning should yield a more precise estimate of
\(\hat{\tau}\). They show that the contribution from each sample unit
(population unit in the sample) to the overall variance is dominated by
local variation. Thus the local neighborhood variance estimator is a
weighted sum of variance estimates from each sample unit's local
neighborhood. These local neighborhoods contain the sample unit itself
and its three nearest neighbors among all other sample units. For more
details, see Stevens and Olsen (2003).

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- for more details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling without
replacement (SRS) and GRTS sampling as well as performance between
design-based (DB) and model-based (MB) inference. In SRS and GRTS
sampling, all population units had equal inclusion probabilities. The
important distinction between SRS and GRTS is that SRS ignores spatial
locations while sampling but GRTS explicitly incorporates them.
Together, the two sampling plans (SRS and GRTS) combined with the two
inference approaches (DB and MB) yielded four sampling-inference
combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB. For SRS-DB, the
Horvitz-Thompson estimator \eqref{eq:ht} was used to estimate means and
the commonly-used SRS variance formula (Lohr, 2009; Särndal et al.,
2003) was used to estimate the variance. This variance formula is given
by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the \(i\)th response value, \(\bar{z}\)
is the mean of all \(z_i\), \(n\) is the sample size, \(N\) is the
population size, and \(f = (1 - n / N)\) (\(f\) is often called the
finite population correction factor). For GRTS-DB, the Horvitz-Thompson
esetimator was used to estimate means and the local neighborhood
variance was used to estimate variances. For SRS-MB and GRTS-MB, FPBK
was used to estimate means and variances and parameters were estimated
using restricted maximum likelihood.

We used simulated data to compare the sampling-inference combinations
across many realized populations from the same data-generating
stochastic process. With the simulated data, we were in control of the
data-generating stochastic process and the random sampling process. We
used real data from the 2012 National Lakes Assessment (USEPA, 2012) to
compare the sampling-inference combinations within a single realized
population (which is typically the case in reality). With the real data,
we were in control of only the random sampling process.

@@

@@

We evaluated performance of the four sampling-inference combinations in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error (DRE). The three sample sizes (\(n\)) were \(n = 50, n = 100,\)
and \(n = 200\). Samples were always selected from a population size
(\(N\)) of \(N = 900\). The two location layouts were random and
gridded. Locations in the random layout were randomly generated inside
the unit square (\([0, 1] \times [0, 1]\)). Locations in the gridded
layout were placed on a fixed, equally spaced grid inside the unit
square. The two response types were normal and skewed. For the normal
response type, the response was simulated using mean-zero random errors
with the exponential covariance (Equation\(~\)\ref{equation:expcov}) for
three proportions of dependent random error (DRE): 0\% DRE, 50\% DRE,
and 90\% DRE. Recall the proportion of DRE is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the DRE variance and independent random error (IRE)
variance from Equation\(~\)\ref{equation:expcov}, respectively. The
total variance, \(\sigma^2_1 + \sigma^2_2\), was always 2. The distance
parameter was always \(\sqrt{2} / 3\), chosen so that the correlation in
the DRE decayed to nearly zero at \(\sqrt{2}\), the largest possible
distance between two population units in the domain. For the skewed
response type, the response was first simulated using the same approach
as for the normal response type, except that the total variance was
0.6931 instead of 2. The response was then exponentiated, yielding a
skewed random variable whose total variance was 2. The skewed responses
were used to evaluate performance of the sampling-inference approaches
for data that were not normal but were still estimated using REML, which
relies on a normal log-likelihood. Figure \ref{fig:sim_pops} shows an
example of a realized population for the normal and skewed responses
using the random layout and 50\% DRE.
